Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-09-28 ms/ms.tex
edited introduction

\keywords{stomatal patterning; Aribidopsis thaliana; amphistomatous
leaves; photosynthesis modeling; optimizing photosynthesis; CO\(_2\)
diffusion}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
physiological responses to the environment regulate gas exchange during
photosynthesis; namely CO\(_2\) assimilation and water loss through
transpiration. Since waxy cuticles are mostly impermeable to CO\(_2\)
and \(H_2O\), stomata are discrete points through which gas exchange
occurs and account only for a small percentage of the leaf area
(\protect\hyperlink{ref-lange_responses_1971}{Lange et al. 1971}). They
consist of two guard cells which open and close upon changes in turgor
pressure or hormonal cues
(\protect\hyperlink{ref-mcadam_linking_2016}{McAdam and Brodribb 2016}).
The stomatal pore leads to an internal space known as the substomatal
cavity where gases (and pathogens) contact the mesophyll
(\protect\hyperlink{ref-muir_stomatal_2020}{Christopher D. Muir 2020}).
Once in the mesophyll, CO\(_2\) diffuses throughout a network of
intercellular air space (IAS) and into mesophyll cells where CO\(_2\)
assimilation (\(A\)) occurs within the chloroplasts
(\protect\hyperlink{ref-lee_diffusion_1964}{Lee and Gates 1964}).
Maximum conductance (\(g_\text{max}\)) and transpiration are determined
by numerous environmental and anatomical parameters such as: vapor
pressure deficit (VPD), irradiance, temperature, wind speed, leaf water
potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Natural selection should optimize the anatomy and physiology of stomata
to maximize CO\(_2\) gain per unit of water loss, thereby maximizing
assimilation rates for a given set of hydraulic constraints and light
environments (\protect\hyperlink{ref-cowan_stomatal_1977}{Cowan and
Farquhar 1977}; \protect\hyperlink{ref-buckley_optimal_2017}{Buckley,
Sack, and Farquhar 2017};
\protect\hyperlink{ref-sperry_predicting_2017}{Sperry et al. 2017}).
However, optimal stomatal anatomy and physiology depends on
developmental and environmental constraints and other physiological
co-variates {[}Croxdale
(\protect\hyperlink{ref-croxdale_stomatal_2000}{2000}); Harrison et al.
(\protect\hyperlink{ref-harrison_influence_2020}{2020}); MUIR IN
PRESS{]}. There is a wide variety of stomatal traits in nature, but
certain common features as well. For example, almost all stomata follow
the one cell spacing rule to maintain proper stomatal functioning

@@

and Blatt 2017}). Stomatal density positively co-varies with irradiance

@@

1987}; \protect\hyperlink{ref-royer_stomatal_2001}{Royer 2001}).
Stomatal size is jointly controlled by genome size, light, and stomatal
density (\protect\hyperlink{ref-jordan_environmental_2015}{Jordan et al.
2015}). Size positively co-varies with genome size

@@

2011}). Total stomatal area (size * density) is optimized for
operational conductance (\(gs_\text{op}\)) rather than maximum
conductance (\(gs_\text{max}\)) such that stomatal apertures are most

@@

\(gs_\text{max}\) (\protect\hyperlink{ref-sack_developmental_2016}{Sack
and Buckley 2016}). Additionally, low stomatal densities lead to
irregular and insufficient CO\(_2\) supply and reduced photosynthetic
efficiency in areas far from stomata
(\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland Pieruschka et
al. 2006}; \protect\hyperlink{ref-morison_lateral_2005}{Morison et al.
2005}), while high stomatal densities can reduce water use efficiency
(WUE) (\protect\hyperlink{ref-bussis_stomatal_2006}{Büssis et al. 2006})
and incur excessive metabolic costs

@@

Amphistomy effectively halves CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer area

@@

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but limitations depend largely
on light availability
(\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott
1990}; \protect\hyperlink{ref-manter_ci_2004}{Manter 2004};
\protect\hyperlink{ref-carriqui_diffusional_2015}{Carriquí et al.
2015}). In low light environments, assimilation rates are low;
consequently, CO\(_2\) flux is low and the internal CO\(_2\)
concentration (\(C_\text{i}\)) remains high; thus \(A_\text{max}\) is
constrained by biochemical limitations in low light
(\protect\hyperlink{ref-kaiser_metabolic_2016}{Kaiser et al. 2016}). For
well hydrated leaves in high light, photosynthesis is often limited by
CO\(_2\) supply as resistances from the boundary layer, stomatal pore,
and mesophyll can result in insufficient \(C_\text{i}\) to supply
efficient photosynthesis at the chloroplast {[}Farquhar, Caemmerer, and
Berry (\protect\hyperlink{ref-farquhar_biochemical_1980}{1980});
lehmeier\_cell\_2017{]}. In this study, we focus primarily on how
stomatal patterning affects diffusion.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length. As the diffusion rate of CO\(_2\) though liquid
is approximately 10,000 times slower than CO\(_2\) diffusion through
air, mesophyll resistance is generally thought to be primarily limited
by liquid diffusion
(\protect\hyperlink{ref-aalto_three-dimensional_2002}{Aalto and Juurola
2002}), but diffusion through the IAS has also been shown to be a rate
limiting process because the tortuous, disjunct nature of the IAS can
greatly increase diffusion path lengths
(\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}). Additionally, tortuosity is higher
in horizontal directions (parallel to leaf surface) than vertical
directions (perpendicular to leaf surface) because of the cylindrical
shape and vertical arrangement of pallisade mesophyll cells
(\protect\hyperlink{ref-earles_beyond_2018}{Earles et al. 2018};
\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}); however, the ratio of lateral to
vertical diffusion rate is still largely unknown
(\protect\hyperlink{ref-morison_lateral_2005}{Morison et al. 2005};
\protect\hyperlink{ref-pieruschka_lateral_2005}{R. Pieruschka 2005};
\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland Pieruschka et al.
2006}). Depending on the thickness of the leaf, porosity of the leaf
mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate
ratio, minimizing diffusion path length for CO\(_2\) via optimally
distributed stomata may yield significant increases in CO\(_2\) supply
for photosynthesis and higher \(A_\text{max}\).

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (Fig. \ref{fig:amphi_grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest
(stomatal zone area: the area a stomate occupies and supplies with
CO\(_2\)). This way, each stomate can be optimally sized for its
relative leaf area and associated CO\(_2\) demand.

To test these hypotheses, we grew \emph{A. thaliana} in high, medium,
and low light and measured stomatal density, size, and patterning on
both leaf surfaces, and spatial coordination between them. We use
voronoi tessellation techniques to calculate the stomatal leaf area. We
also use CO\(_2\) diffusion modeling to ask: what environmental and
physiological conditions maximize the photosynthetic advantage
(\(A\text{gain}\)) of abaxial-adaxial stomatal coordination?
Specifically, we predicted that traits which affect diffusion path
length (leaf thickness, stomatal density, leaf porosity,
lateral-vertical diffusion rate ratio), diffusion rate (temperature,
pressure), and CO\(_2\) demand (Rubisco concentration, light) would
modulate the advantage of optimal stomatal arrangement following the
relationships outlined in Table \ref{tab:hypotheses}. Here, we integrate
over reasonable parameter space to determine the ecophysiological
context most likely to favor stomatal spatial coordination in
amphistomatous leaves.

@@

Finally, to test our hypothesis that stomatal length is modulated by the
area of the leaf to which it supplies CO\(_2\), we examined the
relationship between stomatal zone area and stomatal length using a
Bayesian generalized non-linear multilevel model with the R package
\textbf{brms} version 2.19.0. Stomatal zone area was calculated using
voronoi tessellation (e.g.~Fig. \ref{fig:example}). The stomatal zone
area, \(S_\text{area}\), is the region of the leaf surface whose
distance to stomate, \(S\), is less than the distance to any other
stomate, \(S\). Stomatal length was measured in \texttt{ImageJ}
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-muir_stomatal_2020}{}}%
---------. 2020. {``A {Stomatal} {Model} of {Anatomical} {Tradeoffs}
{Between} {Gas} {Exchange} and {Pathogen} {Colonization}.''}
\emph{Frontiers in Plant Science} 11 (October): 518991.
\url{https://doi.org/10.3389/fpls.2020.518991}.

@@

>>>>

\keywords{amphistomy; Arabidopsis thaliana; CO\(_2\) diffusion; finite
element method; optimality; photosynthesis; stomata}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry points through which gas exchange occurs despite making up a small
percentage of the leaf area
(\protect\hyperlink{ref-lange_responses_1971}{Lange et al. 1971}).
Stomata consist of two guard cells which open and close upon changes in
turgor pressure or hormonal cues
(\protect\hyperlink{ref-mcadam_linking_2016}{McAdam and Brodribb 2016}).
The stomatal pore leads to an internal space known as the substomatal
cavity where gases contact the mesophyll. Once in the mesophyll,
CO\(_2\) diffuses throughout a network of intercellular air space (IAS)
and into mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs
within the chloroplasts (\protect\hyperlink{ref-lee_diffusion_1964}{Lee
and Gates 1964}). Stomatal conductance and transpiration are determined
by numerous environmental and anatomical parameters such as vapor
pressure deficit (VPD), irradiance, temperature, wind speed, leaf water
potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
(\protect\hyperlink{ref-cowan_stomatal_1977}{Cowan and Farquhar 1977};
\protect\hyperlink{ref-buckley_optimal_2017}{Buckley, Sack, and Farquhar
2017}; \protect\hyperlink{ref-sperry_predicting_2017}{Sperry et al.
2017}). However, stomatal anatomy may be partially constrained by
physical and developments limits on phenotypic expression
(\protect\hyperlink{ref-croxdale_stomatal_2000}{Croxdale 2000};
\protect\hyperlink{ref-muir_how_2023}{Christopher D. Muir et al. 2023}).
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning

@@

and Blatt 2017}). Stomatal traits also vary adaptively in different
environments. Stomatal density positively co-varies with irradiance

@@

1987}; \protect\hyperlink{ref-royer_stomatal_2001}{Royer 2001}),
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
(\protect\hyperlink{ref-jordan_environmental_2015}{Jordan et al. 2015}).
Size positively co-varies with genome size

@@

2011}). Total stomatal area (size \(\times\) density) is optimized for
operational conductance (\(gs_\text{s,op}\)) rather than maximum
conductance (\(gs_\text{s,max}\)) such that stomatal apertures are most

@@

\(gs_\text{s,max}\)
(\protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley
2016}), which is proportional to \(gs_\text{s,op}\)
(\protect\hyperlink{ref-murray_consistent_2020}{Murray et al. 2020}).
Additionally, low stomatal densities lead to irregular and insufficient
CO\(_2\) supply and reduced photosynthetic efficiency in areas far from
stomata (\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland
Pieruschka et al. 2006};
\protect\hyperlink{ref-morison_lateral_2005}{Morison et al. 2005}),
while high stomatal densities can reduce water use efficiency (WUE)
(\protect\hyperlink{ref-bussis_stomatal_2006}{Büssis et al. 2006}) and
incur excessive metabolic costs

@@

Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance

@@

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
(\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott
1990}; \protect\hyperlink{ref-manter_ci_2004}{Manter 2004};
\protect\hyperlink{ref-carriqui_diffusional_2015}{Carriquí et al.
2015}). Low light decreasing CO\(_2\) demand by limiting electron
transport rate, leading to relatively high internal CO\(_2\)
concentration (\(C_\text{i}\)) and low \(A_\text{max}\)
(\protect\hyperlink{ref-kaiser_metabolic_2016}{Kaiser et al. 2016}). In
contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis
(\protect\hyperlink{ref-farquhar_biochemical_1980}{Farquhar, Caemmerer,
and Berry 1980}; \protect\hyperlink{ref-lehmeier_cell_2017}{Lehmeier et
al. 2017}). In this study, we focus primarily on how stomatal patterning
affects diffusion, ignoring boundary layer and mesophyll resistances.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length
(\protect\hyperlink{ref-parkhurst_diffusion_1994}{Parkhurst 1994}). As
the diffusion rate of CO\(_2\) though liquid is approximately
\(10^4\times\) slower than CO\(_2\) diffusion through air, mesophyll
resistance is generally thought to be primarily limited by liquid
diffusion (\protect\hyperlink{ref-aalto_three-dimensional_2002}{Aalto
and Juurola 2002}; \protect\hyperlink{ref-evans_resistances_2009}{Evans
et al. 2009}), but diffusion through the IAS has also been shown to be a
rate limiting process because the tortuous, disjunct nature of the IAS
can greatly increase diffusion path lengths
(\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}). Additionally, tortuosity is higher
in horizontal directions (parallel to leaf surface) than vertical
directions (perpendicular to leaf surface) because of the cylindrical
shape and vertical arrangement of pallisade mesophyll cells
(\protect\hyperlink{ref-earles_beyond_2018}{Earles et al. 2018};
\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}). However, the ratio of lateral to
vertical diffusion rate is still largely unknown
(\protect\hyperlink{ref-morison_lateral_2005}{Morison et al. 2005};
\protect\hyperlink{ref-pieruschka_lateral_2005}{R. Pieruschka 2005};
\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland Pieruschka et al.
2006}). Depending on the thickness of the leaf, porosity of the leaf
mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate
ratio, minimizing diffusion path length for CO\(_2\) via optimally
distributed stomata may yield significant increases in CO\(_2\) supply
for photosynthesis and higher \(A_\text{max}\).

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (Fig. \ref{fig:amphi_grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest. We
refer to this as the `stomatal zone', the leaf area surrounding a focal
stomate closest to that stomate and therefore the zone it supplies with
CO\(_2\)). This way, each stomate can be optimally sized relative to the
mesophyll volume it supplies.

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
light) would modulate the advantage of optimal stomatal arrangement
following the relationships outlined in Table \ref{tab:hypotheses}.
Here, we integrate over reasonable parameter space to determine the
ecophysiological context most likely to favor stomatal spatial
coordination in amphistomatous leaves.

@@

Finally, to test our hypothesis that stomatal length is modulated by the
area of the leaf to which it supplies CO\(_2\), we examined the
relationship between stomatal zone area and stomatal length using a
Bayesian generalized non-linear multilevel model with the R package
\textbf{brms} version 2.20.1. Stomatal zone area was calculated using
voronoi tessellation (e.g.~Fig. \ref{fig:example}). The stomatal zone
area, \(S_\text{area}\), is the region of the leaf surface whose
distance to stomate, \(S\), is less than the distance to any other
stomate, \(S\). Stomatal length was measured in \texttt{ImageJ}
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

\leavevmode\vadjust pre{\hypertarget{ref-evans_resistances_2009}{}}%
Evans, J. R., R. Kaldenhoff, B. Genty, and I. Terashima. 2009.
{``Resistances Along the {CO}\(_{\textrm{2}}\) Diffusion Pathway Inside
Leaves.''} \emph{Journal of Experimental Botany} 60 (8): 2235--48.
\url{https://doi.org/10.1093/jxb/erp117}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lehmeier_cell_2017}{}}%
Lehmeier, Christoph, Radoslaw Pajor, Marjorie R. Lundgren, Andrew
Mathers, Jen Sloan, Marion Bauch, Alice Mitchell, et al. 2017. {``Cell
Density and Airspace Patterning in the Leaf Can Be Manipulated to
Increase Leaf Photosynthetic Capacity.''} \emph{The Plant Journal} 92
(6): 981--94. \url{https://doi.org/10.1111/tpj.13727}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-muir_how_2023}{}}%
Muir, Christopher D., Miquel Àngel Conesa, Jeroni Galmés, Varsha S.
Pathare, Patricia Rivera, Rosana López Rodríguez, Teresa Terrazas, and
Dongliang Xiong. 2023. {``How Important Are Functional and Developmental
Constraints on Phenotypic Evolution? {An} Empirical Test with the
Stomatal Anatomy of Flowering Plants.''} \emph{The American Naturalist}
201 (6): 794--812. \url{https://doi.org/10.1086/723780}.

\leavevmode\vadjust pre{\hypertarget{ref-murray_consistent_2020}{}}%
Murray, Michelle, Wuu Kuang Soh, Charilaos Yiotis, Robert A. Spicer,
Tracy Lawson, and Jennifer C. McElwain. 2020. {``Consistent Relationship
Between Field-Measured Stomatal Conductance and Theoretical Maximum
Stomatal Conductance in {C}\(_{\textrm{3}}\) Woody Angiosperms in Four
Major Biomes.''} \emph{International Journal of Plant Sciences} 181 (1):
142--54. \url{https://doi.org/10.1086/706260}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_diffusion_1994}{}}%
---------. 1994. {``Diffusion of {CO}\(_{\textrm{2}}\) and Other Gases
Inside Leaves.''} \emph{New Phytologist} 126 (3): 449--79.
\url{http://www.jstor.org/stable/2557929}.
